Wildcard: Spreadsheet-Driven Customization of Web Applications

### 2020-01-13 paper.tex
final tweaks for paper checklist

\usepackage[backend=biber]{biblatex}
\addbibresource{references.bib}

@@

\author{Geoffrey Litt}
\authorinfo{is a PhD student at MIT in the Software Design Group. His research focuses on programming tools, especially for live programming and end user programming. Contact him at \email{glitt@mit.edu}.}
\author{Daniel Jackson}
\authorinfo{is Professor in the Department of Electrical Engineering and Computer Science at MIT,  a MacVicar teaching fellow, and Associate Director of the Computer Science and Artificial Intelligence Laboratory. He leads the Software Design Group. Contact him at \email{dnj@csail.mit.edu}.}
\affiliation{Massachusetts Institute of Technology}

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Many Web applications do not meet the particular needs of their users.
Browser extensions and user scripts offer a way to customize web
applications, but most people do not have the programming skills to
implement their own extensions.

@@

We have implemented this technique in a prototype browser extension
called Wildcard. Through concrete examples, we demonstrate that Wildcard
can be used to implement useful customizations for real websites.

@@

The user starts by opening the table view, which shows the text and
completed status of each todo, and live updates as the list changes.
They start the customization by adding a new \texttt{snoozeDate} column
to store the snooze date for each todo{ (Figure~\ref{fig:todomvc-demo},
Note A)}.

@@

In order to promote extensibility, Wildcard is implemented as a small
core program along with several types of plugins: site adapters,
formulas, and cell editors. The core contains functionality for
displaying the data table and handling user interactions, and the table
implementation is built using the
\href{https://handsontable.com/}{Handsontable} Javascript library.

\begin{figure}
\hypertarget{fig:architecture}{%
\centering
\includegraphics{media/architecture-clean.png}
\caption{The architecture of the Wildcard
system}\label{fig:architecture}
}
\end{figure}

Site adapters are a key part of Wildcard, since they specify the
bidirectional connection between the web page and its structured data
representation.

Wildcard provides an interface for concisely expressing how the DOM
should be mapped to the spreadsheet table. For example, here is a code
snippet for extracting the name of an Airbnb listing:

@@

Sometimes sophisticated scraping techniques are necessary to extract the
necessary data from the application. For example, we have prototyped a
mechanism for site adapters to observe AJAX requests made by the browser
and extract data directly from JSON responses. This mechanism was used
to implement the Airbnb Walkscore example, since latitude and longitude
aren't shown in the Airbnb UI, but they are available in AJAX responses.
This technique seems promising because AJAX responses tend to already
contain data in a structured form, and web applications increasingly
load data using AJAX. Another technique we might consider adding is the
ability for site adapters to scrape data across multiple pages for
paginated lists of results (as explored in \autocite{huynh2006}).

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter. Instead, Wildcard
automatically mutates the DOM to reflect the spreadsheet state, using
the same declarative spec shown above. The only exception is row actions
(like favoriting an Airbnb listing), which are implemented as imperative
Javascript functions that can can mutate the DOM, simulate clicks on
buttons, etc.

@@

\hypertarget{build-for-an-ecosystem}{%
\subsection{Build for an ecosystem}\label{build-for-an-ecosystem}}

@@

The first party developers of the original software can also play a role
in enabling customization. Spreadsheet-driven customization does not
depend on cooperation from first-party website developers, but if they
were to expose structured data in their web application clients, it
would eliminate the need for third party site adapters. We think there
are compelling reasons for first parties to consider doing this.
Providing Wildcard support would allow users to build extensions to
fulfill their own feature requests. It also would not necessarily
require much effort, since adding Wildcard support could be fairly
straightforward for a first-party that already has direct access to the
structured data in the page. There is also precedent for first parties
implementing an official client extension API: for several years, Google
maintained an official extension API in Gmail for Greasemonkey scripts
to use.\footnote{Incidentally, since then, third parties have continued
  to maintain stable Gmail extension APIs used by many browser
  extensions \autocite{streak,talwar2019}, illustrating the potential of
  collaboratively maintaining third party adapters.}

@@

In the broadest sense, Wildcard is inspired by systems aiming to make
software into a dynamic medium where end users frequently create and
modify software to meet their own needs, rather than only consuming
applications built by programmers. These systems include Smalltalk
\autocite{kay1977}, Hypercard \autocite{hypercard2019} , Boxer
\autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
Dynamicland \autocite{victor}.\footnote{The project's name Wildcard
  comes from the internal pre-release name for Hypercard, which doubly
  inspired our work by promoting both software modification by end users
  and the ideas behind the Web.}

>>>>

\usepackage[backend=biber]{biblatex}
\addbibresource{references2.bib}

@@

\author{Geoffrey Litt}
\authorinfo{is a PhD student at MIT in the Software Design Group. His research focuses on programming languages and environments, particularly tools for end user programming and live programming. Contact him at \email{glitt@mit.edu}.}
\author{Daniel Jackson}
\authorinfo{is Professor in the Department of Electrical Engineering and Computer Science at MIT,  a MacVicar teaching fellow, and Associate Director of the Computer Science and Artificial Intelligence Laboratory. He leads the Software Design Group. Contact him at \email{dnj@csail.mit.edu}.}
\affiliation{Massachusetts Institute of Technology}

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Many Web applications do not meet the particular needs of their users.
Browser extensions offer a way to customize web applications, but most
people do not have the programming skills to implement their own
extensions.

@@

We have implemented this technique in a prototype browser extension
called Wildcard. Through concrete examples, we demonstrate that Wildcard
can support useful customizations---ranging from sorting lists of search
results to showing related data from web APIs---on top of existing
websites. We also present the design principles underlying the
prototype.

Enabling easier customization can dramatically improve people's
experiences with software. We think that spreadsheet-driven
customization offers a promising new approach to the problem.

@@

The user opens the table view, which shows the text and completed status
of each todo, and updates live in response to changes. They start the
customization by adding a new \texttt{snoozeDate} column to store the
snooze date for each todo{ (Figure~\ref{fig:todomvc-demo}, Note A)}.

@@

In order to promote extensibility, Wildcard is implemented as a small
core program along with several types of plugins: site adapters,
formulas, and cell editors. The core contains functionality for
displaying the data table and handling user interactions, and is built
on the \href{https://handsontable.com/}{Handsontable} Javascript
library.

\begin{figure}
\hypertarget{fig:architecture}{%
\centering
\includegraphics{media/architecture-clean.png}
\caption{The architecture of the Wildcard
prototype}\label{fig:architecture}
}
\end{figure}

Site adapters are a key part of Wildcard, since they specify the
bidirectional connection between the web page and its structured data
representation. Wildcard represents data in a relational table with
named and typed columns, and provides an interface for concisely
expressing how the DOM should be mapped to this structure. For example,
here is a code snippet for extracting the name of an Airbnb listing.

@@

Sometimes, more advanced scraping techniques are necessary to extract
the necessary data from the application. For example, we have prototyped
a mechanism for site adapters to observe AJAX requests made by the
browser and extract data directly from JSON responses. This mechanism
was used to implement the Airbnb Walkscore example, since latitude and
longitude aren't shown in the Airbnb UI, but they are available in AJAX
responses. This technique seems promising because AJAX responses already
contain data in a structured form, and web applications are increasingly
loading data using AJAX. Another technique we might consider is enabling
site adapters to scrape data across multiple pages for paginated lists
of results (as explored in \autocite{huynh2006}).

Because Wildcard requires a live bidirectional connection between the UI
and the structured data, there are some additional challenges beyond
those typically associated with web scraping.

First, the site adapter also needs to support sending updates from the
table to the original page. We have made this fairly straightforward,
because most DOM manipulation is not performed directly by the site
adapter---instead, Wildcard automatically mutates the DOM to reflect the
spreadsheet state, using the same declarative spec shown above. The only
exception to this is row actions (like favoriting an Airbnb listing),
which are simply implemented as Javascript functions with the freedom to
perform arbitrary behaviors like simulating clicks on buttons and
mutating the DOM.

Another challenge is triggering updates to the spreadsheet data in
response to UI changes that happen after initial page load. Site
adapters are responsible for implementing site-specific logic for
recognizing these changes only by observing the DOM. So far, we have
been able to use event listeners and the MutationObserver API to
successfully observe changes, but it may prove challenging to correctly
respond to updates on some sites only by observing the DOM.

@@

\hypertarget{design-for-an-ecosystem-of-users}{%
\subsection{Design for an ecosystem of
users}\label{design-for-an-ecosystem-of-users}}

@@

The first party developers of the original software can also play a role
in enabling customization. Spreadsheet-driven customization does not
depend on cooperation from first-party website developers, but if they
were to expose structured data in their web application clients, it
would eliminate the need for third party site adapters. We think there
are compelling reasons for first parties to consider doing this.
Providing Wildcard support would allow users to build extensions to
fulfill their own feature requests. It also would not necessarily
require much effort, since adding Wildcard support could be fairly
straightforward for a first-party that already has direct access to the
structured data in the page. There is also precedent for first parties
implementing an official client extension API: for several years, Google
maintained an official extension API in Gmail for Greasemonkey scripts
to use.

Google no longer maintains its own API, but third parties have continued
to maintain successful replacements---Gmail.js \autocite{talwar2019} is
an open source wrapper API with 77 contributors and hundreds of commits.
This shows another benefit of designing for an ecosystem: it's easier
for many programmers to collectively maintain a single adapter for a
site and keep it updated in response to changes, instead of creating
custom scraping logic for individual extensions.

@@

In the broadest sense, Wildcard is inspired by systems aiming to make
software into a dynamic medium where end users frequently create and
modify software to meet their own needs, rather than only consuming
applications built by programmers. These systems include Smalltalk
\autocite{kay1977}, Hypercard \autocite{hypercard2019} , Boxer
\autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
Dynamicland \autocite{victor}. (The project's name Wildcard comes from
the internal pre-release name for Hypercard, which doubly inspired our
work by promoting both software modification by end users and the ideas
behind the Web.)
